Testing the correlates of vast vertebrae number disparity in Australian blindsnakes

### 2024-10-11 manuscript.Rmd
update before sending to co-authors

---
# title: "Dissociation between number of vertebrae and body size in Australian blindsnakes"
title: "Testing the correlates of huge vertebrae number disparity in Australian blindsnakes"
date: "Version `r Sys.Date()`"
author: ""
output: 
  # bookdown::html_document2:
  #   toc: true
  #   number_sections: FALSE
  bookdown::word_document2:
    reference_docx: [./custom-reference-doc.docx]
    number_sections: false
    fig_caption: yes
bibliography: [C:/Users/ST/Documents/repo/anu_thesis_texts/zotero_library.bib]
csl: evolution.csl
link-citations: true
editor_options:
  chunk_output_type: console
---

@@

The evolution of the axial skeleton is important in understanding the evolution of body plans in vertebrates. Changes in the vertebral skeleton can influence body form and modes of locomotion. Elongate animals are commonly show a positive correlation between body size and the number of vertebrae, a condition known as pleomerism. However, some groups of vertebrates show a lack of pleomerism, suggesting it is possible to dissociate the production of segment and somatic growth, but what factors influenced this pattern is not well understood. Snakes offer a great model to study axial evolution because the number of vertebrae vary greatly between species and they exhibit a wide range of body size and shape. Here, we investigate the prevalence of pleomerism in a continental radiation of fossorial blindsnakes and test whether body shape and ecology can explain the variation in vertebrae number. We obtained X-ray scans and morphometric data for 42 described species of *Anilios* blindsnakes and evaluated their correlation with each other and ecological variables. We found an absence of pleomerism among Australian blindsnakes and a correlation between vertebrae ratio and body shape, where narrow-and-long species usually possess more vertebrae for their maximum length than wide-and-robust species. Finally, we found that species that have a greater number of vertebrae for their size are generally found in warmer environments and environments with harder-to-penetrate soil substrate. Our results extend our understanding of the axial evolution of basal snakes and the macroevolutionary dynamics of Australian blindsnakes.

@@

The lack of pleomerism in some species suggests it is possible to disassociate segment production with body growth, and thus reveals the developmental processes underlying evolutionary patterns. Another clue to disassociation in reptiles can be found in the relationship between the number of ventral scales and precloacal vertebrae – a correlation that is 1:1 in many snakes [@alexanderPattern1966] and so ventral scales have often been used as a proxy for vertebrae number [@lindellEvolution1994;@shineVertebral2000]. However, this 1:1 correlation is absent in file snakes, *Achrochodus*, and some blindsnakes, Typhlopidae, [@alexanderPattern1966;@headDissociation2007], as well as some sea snakes of the genus *Hydrophis* [@vorisDermal1975], suggesting these taxa do not conform to pleomerism.
Snakes are therefore a good model to study axial evolution because they have great diversity in vertebrae number and a wide range of adult body sizes and shapes but without a limb skeleton to confound signals of locomotor adaptation. Recent studies have indicated the number of vertebrae varies predictably with ecological diversification in elapid snakes [@sherrattVertebral2019;@hamptonInterspecific2019;@sherrattMacroevolution2022] and colubrid snakes [@nash-hahnModification2024]. 
The number of vertebrae has also been shown to influence snake locomotion, whereby the number of vertebrae negatively correlates with speed travelled in garter snakes [@kelleyEffects1997]. Despite these recent advances, there have been no study that have quantified the number of vertebrae across all the species in a diverse assemblage of fossorial snakes to test for the variation and potential ecological drivers.

@@

The relationship between segmentation and body elongation in scolecophidians is not well understood. However, a study by Head and Polly [-@headDissociation2007] that included 49 species across two genera typhlopids showed that vertebrae number and body size are positively correlated except when some of the largest species were included in the analysis. In particular, the largest species of blindsnakes in the African continent (*Afrotyphlops schlegelii*) have more precloacal vertebrae than expected for their size [@headDissociation2007]. This suggests that elongation in these species is achieved by elongating each vertebra rather than adding additional vertebrae. Compared to other snakes, the vertebral column of fossorial snakes tends to feature a much higher proportion of pre- than post-cloacal segments [@pollyTesting2001], which is reflected in the relatively short tails of blindsnakes [@tiatragulMorphological2024;@szyndlarIllustrated2023]. Therefore, blindsnakes are a good system to test how changes in the number of pre-cloacal vertebrae affects maximum body length.

@@

The Australian blindsnake radiation, *Anilios*, comprise 47 described species distributed across all major biomes of Australia [@tiatragulAustralia2023;@tiatragulPaleoenvironmental2023]. Species in the genus show considerably diverse morphology from narrow-and-long to robust-and-wide [@tiatragulMorphological2024;@ehmannFamily1993]. However, information about the relationship between body size and vertebrae number is unknown. In this study, we investigate the prevalence of pleomerism among Australian blindsnakes and test hypotheses about how deviations from pleomerism may be explained by different body shapes or other ecological factors among Australian typhlopid snakes. The specific questions we addressed are: (1) How prevalent is pleomerism among Australian blindsnakes? (2) Can deviations from pleomerism be explained by body shape in Australian blindsnakes? (3) Does vertebral ratio vary with their ecology (e.g., soil compactness and mean annual temperature)?

@@

We found a dissociation between vertebrae number and maximum adult body length in Australian blindsnake. While pleomerism has been documented in most snake groups [@lindellEvolution1994], exceptionally large species ("giant taxa") within boids, pythonids, and blindsnakes such as *Afrotyphlops schlegelii* have fewer than expected pre-cloacal vertebrae for their size [@headDissociation2007]. In line with these observations, some of the largest species in our dataset (*A. torresianus* and *A. pinguis*) possess lower than average number of vertebrae (Fig. \@ref(fig:fig-tbl-vert)). The absence of pleomerism suggest that the large size attained by these "giant taxa" may be due to modification of post-somitogenetic somatic growth [@headDissociation2007]. In other words, species with larger body size do not simply have more vertebrae segments, rather the size of each vertebral segment has been elongated during growth. Several mechanisms have been suggested for how giant amniotes achieve their large size including prolonging their growth [@ericksonHow1999] and increasing the rate of growth [@ericksonGigantism2004]. Unfortunately, without life-history data for blindsnakes, we cannot test if these mechanisms may be responsible. However, elongation may also be achieved in other ways.
The regionalisation of vertebrae, where distinct regions along the axial column exhibit morphological diversification, and evolve relatively independently, may explain the lack of pleomerism in blindsnakes. For example, the absence of pleomerism in *Hydrophis* sea snakes has been attributed to the increase in the number of smaller vertebrae in the forebody region compared to larger vertebrae in the hindbody region [@sherrattMacroevolution2022;@sherrattVertebral2019]. Similarly, colubrids have also been shown to have different regionalisation patterns in the vertebral column that contributes to elongation in different ways [@nash-hahnModification2024]. 
The position of the heart is a key marker for delimiting the regions of the snake axial column [@hamptonHeart2022;@hamptonRegionalization2023]. Future research should investigate the regionalisation of the vertebral column in blindsnakes as this will provide important insights to understanding the lack of pleomerism.

The lack of pleomerism among burrowing vertebrates is not uncommon. @lindellEvolution1994 found that among Australian elapids and African snakes, burrowing species had fewer vertebrae than expected from their body length compared to terrestrial and arboreal species. The dissociation between vertebrae number and body size has also been documented in caecilians, a primarily fossorial order of amphibians. In caecilians, @renousBody1989 found that members of the same genus will vary in size but generally share similar number of vertebrae, indicating a lack of pleomerism except for in a few genera such as *Oscaecilia*. Another group of highly specialised burrowing reptile is the amphisbaenians [@gansCharacteristics1978;@gansAmphisbaeniansreptiles1969]. While several studies have documented the number of vertebrae in some species [@bergmannVertebral2012;@gansStudies1960;@gansStudies1971;@gansCharacteristics1978;@alexanderPattern1966], no study has investigated the correlation between vertebrae number and body size in amphisbaenians. However, they have been shown to deviate from the expected 1:1 ratio of ventral scale number to vertebrae [@alexanderPattern1966], suggesting a lack of pleomerism. We recommend future research to comprehensively investigate the prevalence of pleomerism among burrowing limbless and limb-reduced vertebrates to improve our understanding of the evolution of axial system.

@@

Aspect ratio (body shape) varied with the number of vertebrae in similar ways as previously reported in fish and other snakes (e.g. vipers and colubrids), where species with a more elongated body had relatively more vertebrae than stouter ones [@lindseyPleomerism1975;@lindellEvolution1994;@wardAxial2010]. Two possible explanations have been proposed by @lindellEvolution1994 for snakes including adaptations for locomotion and prey consumption, which are mirrored in the interpretations from elongate fish [@wardAxial2010]. Regarding locomotion, @poughSpecializations1983 showed that in snakes, longer body segments may be advantageous for elongated species that usually move faster, whereas shorter segments are favoured for stouter species that move more slowly. In blindsnakes, however, movement speed may not be as important as how much push force they can generate when burrowing. Snakes with relatively wider bodies are generally able to produce greater push force for their body size [@herrelBurrowing2021], while species with narrower bodies will be able to use pre-existing small crevices and galleries without needing to expend any energy. In regards to prey consumption, stout species with large heads (i.e. vipers) tend to consume large bulky prey, which could be facilitated by having a stout body as this minimises disruption of physiological functions [@poughSpecializations1983;@lindellEvolution1994;@moonFeeding2019]. Although Australian blindsnakes primarily prey on pupae and larvae of ants and termites, there is some variation in the size of prey consumed [@webbDietary1993]. In particular, heavy-bodied blindsnake species (e.g. *A. pinguis*, *A. proximus*, and *A. ligatus*) fed on larger species of ants compared to more slender-bodied blindsnakes species (e.g. *A. hamatus*, *A. bituberculatus*, *A. waitii*, *A. weidii*) [@webbDietary1993].

>>>>

---
# title: "Dissociation between number of vertebrae and body size in Australian blindsnakes"
title: "Testing the correlates of vast vertebrae number disparity in Australian blindsnakes"
date: "Version `r Sys.Date()`"
author: ""
output: 
  # bookdown::html_document2:
  #   toc: true
  #   number_sections: FALSE
  bookdown::word_document2:
    reference_docx: [./custom-reference-doc.docx]
    number_sections: false
    fig_caption: yes
bibliography: [C:/Users/ST/Documents/repo/anu_thesis_texts/zotero_library.bib]
csl: evolution.csl
link-citations: true
editor_options:
  chunk_output_type: console
---

@@

The evolution of the axial skeleton is important in understanding the evolution of body plans in vertebrates. Changes in the vertebral skeleton can influence body form and modes of locomotion. Elongate animals are commonly show a positive correlation between body size and the number of vertebrae, a condition known as pleomerism. However, some groups of vertebrates show a lack of pleomerism, suggesting it is possible to dissociate the production of segment and somatic growth, but what factors influenced this pattern is not well understood. Snakes offer a great model to study axial evolution because the number of vertebrae vary greatly between species and they exhibit a wide range of body size and shape. Here, we investigate the prevalence of pleomerism in a continental radiation of fossorial blindsnakes and test whether body shape and ecology can explain the variation in vertebrae number. We obtained X-ray scans and morphometric data for 42 described species of *Anilios* blindsnakes and evaluated their correlation with each other and ecological variables. We found an absence of pleomerism among Australian blindsnakes and a correlation between vertebrae ratio and body shape, where narrow-and-long species usually possess more vertebrae for their maximum length than wide-and-robust species. We also found that species that have a greater number of vertebrae for their size are generally found in warmer environments and environments with harder-to-penetrate soil substrate. Our results extend our understanding of the axial evolution of basal snakes and the macroevolutionary dynamics of Australian blindsnakes.

@@

The lack of pleomerism in some species suggests it is possible to disassociate segment production with body growth, and thus reveals the developmental processes underlying evolutionary patterns. Another clue to disassociation in reptiles can be found in the relationship between the number of ventral scales and precloacal vertebrae – a correlation that is 1:1 in many snakes [@alexanderPattern1966] and so ventral scales have often been used as a proxy for vertebrae number [@lindellEvolution1994;@shineVertebral2000]. However, this 1:1 correlation is absent in file snakes, *Achrochodus*, and some blindsnakes, Typhlopidae, [@alexanderPattern1966;@headDissociation2007], as well as some sea snakes of the genus *Hydrophis* [@vorisDermal1975], suggesting these taxa do not conform to pleomerism.
Snakes are therefore a good model to study axial evolution because they have great diversity in vertebrae number and a wide range of adult body sizes and shapes but without a limb skeleton to confound signals of locomotor adaptation. Recent studies have indicated the number of vertebrae varies predictably with ecological diversification in elapid snakes [@sherrattVertebral2019;@hamptonInterspecific2019;@sherrattMacroevolution2022] and colubrid snakes [@nash-hahnModification2024]. 
The number of vertebrae has also been shown to influence snake locomotion, whereby the number of vertebrae negatively correlates with speed travelled in garter snakes [@kelleyEffects1997]. Despite these recent advances, there have been no studies that have quantified the number of vertebrae across all the species in a diverse assemblage of fossorial snakes to test for the variation and potential ecological drivers.

@@

The relationship between segmentation and body elongation in scolecophidians is not well understood. However, a study by Head and Polly [-@headDissociation2007] that included 49 species across two typhlopids genera showed that vertebrae number and body size are positively correlated except when some of the largest species were included in the analysis. In particular, the largest species of blindsnakes on the African continent (*Afrotyphlops schlegelii*) have more precloacal vertebrae than expected for their size [@headDissociation2007]. This suggests that elongation in these species is achieved by elongating each vertebra rather than adding additional vertebrae. Compared to other snakes, the vertebral column of fossorial snakes tends to feature a much higher proportion of pre- than post-cloacal segments [@pollyTesting2001], which is reflected in the relatively short tails of blindsnakes [@tiatragulMorphological2024;@szyndlarIllustrated2023]. Therefore, blindsnakes are a good system to test how changes in the number of pre-cloacal vertebrae affects maximum body length.

@@

The Australian blindsnake radiation, *Anilios*, comprise 47 described species distributed across all major biomes of Australia [@tiatragulAustralia2023;@tiatragulPaleoenvironmental2023]. This radiation displays considerable morphological diversity which ranges from narrow-and-long to robust-and-wide species [@tiatragulMorphological2024;@ehmannFamily1993]. However, information about the relationship between body size and vertebrae number is unknown. In this study, we investigate the prevalence of pleomerism among Australian blindsnakes and test hypotheses about how deviations from pleomerism may be explained by different body shapes or other ecological factors among Australian typhlopid snakes. The specific questions we addressed are: (1) How prevalent is pleomerism among Australian blindsnakes? (2) Can deviations from pleomerism be explained by body shape in Australian blindsnakes? (3) Does vertebral ratio vary with their ecology (e.g., soil compactness and mean annual temperature)?

@@

We found a dissociation between vertebrae number and maximum adult body length in Australian blindsnake. While pleomerism has been documented in most snake groups [@lindellEvolution1994], exceptionally large species ("giant taxa") within boids, pythonids, and blindsnakes such as *Afrotyphlops schlegelii* have fewer than expected pre-cloacal vertebrae for their size [@headDissociation2007]. In line with these observations, some of the largest species in our dataset (*A. torresianus* and *A. pinguis*) possess lower than average number of vertebrae (Fig. \@ref(fig:fig-tbl-vert)). The absence of pleomerism suggests that the large size attained by these "giant taxa" may be due to modification of post-somitogenetic somatic growth [@headDissociation2007]. In other words, species with larger body size do not simply have more vertebrae segments, rather the size of each vertebral segment has been elongated during growth. Several mechanisms have been suggested for how giant amniotes achieve their large size including prolonging their growth [@ericksonHow1999] and increasing the rate of growth [@ericksonGigantism2004]. Unfortunately, without life-history data for blindsnakes, we cannot test if these mechanisms may be responsible. However, elongation may also be achieved in other ways.
The regionalisation of vertebrae, where distinct regions along the axial column exhibit morphological diversification, and evolve relatively independently, may explain the lack of pleomerism in blindsnakes. For example, the absence of pleomerism in *Hydrophis* sea snakes has been attributed to the increase in the number of smaller vertebrae in the forebody region compared to larger vertebrae in the hindbody region [@sherrattMacroevolution2022;@sherrattVertebral2019]. Similarly, colubrids have also been shown to have different regionalisation patterns in the vertebral column that contributes to elongation in different ways [@nash-hahnModification2024]. 
The position of the heart is a key marker for delimiting the regions of the snake axial column [@hamptonHeart2022;@hamptonRegionalization2023]. Future research should investigate the regionalisation of the vertebral column in blindsnakes as this will provide important insights to understanding the lack of pleomerism.

The lack of pleomerism among burrowing vertebrates is not uncommon. @lindellEvolution1994 found that among Australian elapids and African snakes, burrowing species had fewer vertebrae than expected from their body length compared to terrestrial and arboreal species. The dissociation between vertebrae number and body size has also been documented in caecilians, a primarily fossorial order of amphibians. In caecilians, @renousBody1989 found that members of the same genus will vary in size but generally share similar number of vertebrae, indicating a lack of pleomerism except for in a few genera such as *Oscaecilia*. 
While several studies have documented the number of vertebrae in the highly specialised burrowing amphisbaenianss [@bergmannVertebral2012; @gansAmphisbaeniansreptiles1969; @gansStudies1960;@gansStudies1971;@gansCharacteristics1978;@alexanderPattern1966], no study has investigated the correlation between vertebrae number and body size this group. However, they have been shown to deviate from the expected 1:1 ratio of ventral scale number to vertebrae [@alexanderPattern1966], suggesting a lack of pleomerism. We recommend future research to comprehensively investigate the prevalence of pleomerism among burrowing limbless and limb-reduced vertebrates to improve our understanding of the evolution of axial system.

@@

Aspect ratio (body shape) varied with the number of vertebrae in similar ways as previously reported in fish and other snakes (e.g. vipers and colubrids), where species with a more elongated body had relatively more vertebrae than stouter ones [@lindseyPleomerism1975;@lindellEvolution1994;@wardAxial2010]. Two possible explanations have been proposed by @lindellEvolution1994 for snakes including adaptations for locomotion and prey consumption, which are mirrored in the interpretations from elongate fish [@wardAxial2010]. Regarding locomotion, @poughSpecializations1983 showed that in snakes, longer body segments may be advantageous for elongated species that usually move faster, whereas shorter segments are favoured for stouter species that move more slowly. In blindsnakes, however, movement speed may not be as important as how much push force they can generate when burrowing. Snakes with relatively wider bodies are generally able to produce greater push force for their body size [@herrelBurrowing2021], while species with narrower bodies will be able to use pre-existing small crevices and galleries without needing to expend any energy. In regard to prey consumption, @poughSpecializations1983 suggest that stout species with large heads (i.e. vipers) tend to consume large bulky prey, which could be facilitated by having a stout body as this minimises disruption of physiological functions. However, Australian blindsnakes primarily prey on pupae and larvae of ants and termites, which are relatively small [@webbDietary1993]. As such, we do not expect prey consumption to influence the relationhship between aspect ratio and vertebrae we observed. 
<!-- In particular, heavy-bodied blindsnake species (e.g. *A. pinguis*, *A. proximus*, and *A. ligatus*) fed on larger species of ants compared to more slender-bodied blindsnakes species (e.g. *A. hamatus*, *A. bituberculatus*, *A. waitii*, *A. weidii*).  -->
